---
title: "BAM_reads_filtering"
author: "Dani"
date: "14 de enero de 2019"
output: html_document
---

#0: Define paths.

```{r Define paths, eval=FALSE, engine='bash'}

S_PATH=/opt/snpEff #software path
C_PATH=/home/dkleinman/datos/snpEff #config file path
O_PATH=/home/dkleinman/datos/snpEff #output path
I_PATH=/home/GRUPOS/grupolince/immunocapture/prueba_highdiv #immunocapture path
V_PATH=/GRUPOS/grupolince/lynx_genomes_5x/VCFs_Dani/nmVCFs #VCFs path
G_PATH=/GRUPOS/grupolince/lynx_genomes_5x/gVCFs #gVCFs path
B_PATH=/home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final #BAM files path
REF=/home/GRUPOS/grupolince/reference_genomes/lynx_pardinus_genome/lp23.fa #path to reference genome
GATK=/opt/GATK-3.7/GenomeAnalysisTK.jar #GATK software path
BCF=/opt/bcftools-1.6/bcftools #BCFtools software path

```
#3: Perform NM-based BAM filterings.

##Explore the contamination.
###Filter in reads with NM ≥ 12.

```{r Perform NM-based BAM filterings, eval=FALSE, engine='bash'}

cd /home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final/BAM_genes_5x
screen -S c_lp_do_genes_hm_BAMs.log
script c_lp_do_genes_hm_BAMs.log

cd /home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final/BAM_genes_5x
declare SAMPLES=$(ls c_lp_do*.genes.bam | cut -c9-12 | sort | uniq)
cd /opt/bamtools/lib
for i in ${SAMPLES[@]}
  do
  echo "${i}"
  bamtools filter -tag "NM:>12" -in /home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final/BAM_genes_5x/c_lp_do_"${i}"_recal_round-1.genes.bam -out /home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final/BAM_genes_5x/c_lp_do_"${i}"_recal_round-1.genes-hm.bam
  samtools index c_lp_do_"${i}"_recal_round-1.genes-hm.bam
  done

```

###Convert reads to FASTA.

```{r Perform NM-based BAM filterings, eval=FALSE, engine='bash'}

cd /home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final/BAM_genes_5x
screen -S c_lp_do_genes_hm_FASTAs.log
script c_lp_do_genes_hm_FASTAs.log

cd /home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final/BAM_genes_5x
declare SAMPLES=$(ls c_lp_do*genes-hm.bam | cut -c9-12 | sort | uniq)
for i in ${SAMPLES[@]}
  do
  echo "${i}"
  samtools fasta c_lp_do_"${i}"_recal_round-1.genes-hm.bam > c_lp_do_"${i}"_recal_round-1.genes-hm.fa
  done

```

###Filter in reads with NM ≥ 5 and convert to fasta.

```{r Perform NM-based BAM filterings, eval=FALSE, engine='bash'}

cd /home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final/BAM_genes_5x
screen -S c_lp_do_genes_hm5_BAMs.log
script c_lp_do_genes_hm5_BAMs.log

cd /home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final/BAM_genes_5x
declare SAMPLES=$(ls *.genes.bam | cut -c1-12 | sort | uniq)
for i in ${SAMPLES[@]}
  do
  echo "${i}"
  bamtools filter -tag "NM:>5" -in "${i}"_recal_round-1.genes.bam -out "${i}"_recal_round-1.genes-hm5.bam
  samtools index "${i}"_recal_round-1.genes-hm5.bam
  samtools fasta "${i}"_recal_round-1.genes-hm5.bam > "${i}"_recal_round-1.genes-hm5.fa
  done

```

##Filter out reads with high NM. Based on the previously obtained distribution of NM, we decide to filter out reads with NM>4.
###Genes-only BAMs.
####DON:
```{r Perform NM-based BAM filterings, eval=FALSE, engine='bash'}

cd /home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final/BAM_genes_5x
screen -S c_lp_do_genes_nm_BAMs.log
script c_lp_do_genes_nm_BAMs.log

cd /home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final/BAM_genes_5x
declare SAMPLES=$(ls c_lp_do_*_recal_round-1.genes.bam | cut -c9-12 | sort | uniq)
for i in ${SAMPLES[@]}
  do
  echo "${i}"
  bamtools filter -tag "NM:<=4" -in c_lp_do_"${i}"_recal_round-1.genes.bam -out c_lp_do_"${i}"_recal_round-1.genes-nm.bam
  samtools index c_lp_do_"${i}"_recal_round-1.genes-nm.bam
  done

```




#4: Perform variant calling. Combine all BAMs of interest into the separate per species VCF (these won't include substitutions between species but their variants will be more accurate) or into a combined VCF (to track substitutions).

##B: Whole-genome. 
###For both together.

```{r Perform variant calling, eval=FALSE, engine='bash'}

#Perform direct variant calling (without gVCFs) on the NM≤4 filtered Lynx pardinus BAMs.
cd $V_PATH
screen -S c_lp_sm_c_lp_do_c_ll_ki_c_ll_no_c_ll_po_nm.log
script c_lp_sm_c_lp_do_c_ll_ki_c_ll_no_c_ll_po_nm.log

cd $B_PATH
java -XX:MaxMetaspaceSize=1g -XX:+UseG1GC -XX:+UseStringDeduplication -jar $GATK \
-T HaplotypeCaller \
-R $REF \
$(for var in c_*recal_round-1.nm.bam; do echo -I ${var}" ";done) \
-o $V_PATH/c_lp_sm_c_lp_do_c_ll_ki_c_ll_no_c_ll_po_nm.vcf

grep -v '#' $V_PATH/c_lp_sm_c_lp_do_c_ll_ki_c_ll_no_c_ll_po_nm.vcf | wc -l #1341848

#Rename the samples with wrong names.
cd $V_PATH
bcftools query -l c_lp_sm_c_lp_do_c_ll_ki_c_ll_no_c_ll_po_nm.vcf #check sample names
#!/bin/bash
#cat << "EOF" > lp_ll_rename.txt #Unmark this line when running it (I marked it in R because it doesn't understand it and messes the colours of subsequent lines). It should include all incorrect names.
h_lp_do_0007 c_lp_do_0007
LL90_rgsm c_ll_ki_0090
EOF
cat lp_ll_rename.txt
bcftools reheader -s lp_ll_rename.txt -o c_lp_sm_c_lp_do_c_ll_ki_c_ll_no_c_ll_po_nm_renamed.vcf c_lp_sm_c_lp_do_c_ll_ki_c_ll_no_c_ll_po_nm.vcf #copy the VCF and rename the wrong named samples in the new VCF
rm lp_ll_rename.txt
mv c_lp_sm_c_lp_do_c_ll_ki_c_ll_no_c_ll_po_nm_renamed.vcf c_lp_sm_c_lp_do_c_ll_ki_c_ll_no_c_ll_po_nm.vcf

grep -v '#' c_lp_sm_c_lp_do_c_ll_ki_c_ll_no_c_ll_po_nm.vcf | wc -l #

```

#5: Polarize contemporary VCFs. Use VCFtools in order to polarize (ancestral vs. derived) the contemporary ll & lp VCFs. The ancestral state was inferred by Maria.

##Prepare ancestral genome fasta.(This is already done).

```{r Polarize contemporary VCF, eval=FALSE, engine='bash'}

#This is the command to use in order to add to a VCF information on the ancestral/derived alleles: /opt/vcftools_0.1.13/perl/fill-aa. There's documentation inside that command that I'll follow here.

#First, the ancestral alleles file should be bgzipped (according to the documentation, they should be gzipped, but later on when trying to run faidx I got an error stating that files should be gzipped in order to build a fai index):
bgzip -c /home/GRUPOS/grupolince/reference_genomes/lynx_rufus_genome/c_lr_zz_0001_recal1.fa > /home/GRUPOS/grupolince/reference_genomes/lynx_rufus_genome/c_lr_zz_0001_recal1.fa.gz

#Next they should be fai indexed:
/opt/samtools-1.6/samtools faidx /home/GRUPOS/grupolince/reference_genomes/lynx_rufus_genome/c_lr_zz_0001_recal1.fa.gz

```

##Use vcftools to add Ancestral Allele annotation to the VCF.

```{r Polarize contemporary VCF, eval=FALSE, engine='bash'}

#Code to annotate the AA (ancestral alleles) subfield from the INFO field in the VCF. I keep getting the following error: "Can't locate Vcf.pm in @INC". I look it up and apparently it's easy to solve: an environment variable PERL5LIB should be defined as the path to perl.
cd $V_PATH
screen -S c_lp_sm_c_lp_do_nm_genes_aafilled.log
script c_lp_sm_c_lp_do_nm_genes_aafilled.log

export PERL5LIB=/opt/vcftools_0.1.13/perl/ #set required environmental variable
V_PATH=/GRUPOS/grupolince/lynx_genomes_5x/VCFs_Dani/nmVCFs #VCFs path

cat $V_PATH/c_lp_sm_c_lp_do_nm_genes.vcf | /opt/vcftools_0.1.13/perl/fill-aa -a /home/GRUPOS/grupolince/reference_genomes/lynx_rufus_genome/c_lr_zz_0001_recal1.fa.gz | bgzip -c > c_lp_sm_c_lp_do_nm_genes_aafilled.vcf.gz #I tried to run this first without bgzipping it, and it was increasingly slow. Each day it did around half the previous day. When the server crashed, I relaunched it adding the bgzip command and it finished in a little over 24h.

gzip -d -c c_lp_sm_c_lp_do_nm_genes_aafilled.vcf.gz > c_lp_sm_c_lp_do_nm_genes_aafilled.vcf #Get decompressed version. I ran md5sum for both the gzipped and the unzipped versions and there were no differences.

grep -v '#' c_lp_sm_c_lp_do_nm_genes_aafilled.vcf | wc -l #1341848

```

##Use VcfFilterJdk to polarize the AA-filled VCF.

```{r Polarize contemporary VCF, eval=FALSE, engine='bash'}

#Polarize the AA-filled VCF based on the new INFO/AA column. Alleles will be switched whenever the ancestral allele matches the alternative one, and genotypes will be properly recoded as well. The following code was originally provided by Pierre Lindenbaum and modified by José Luis Castro.

cd $V_PATH
screen -S c_lp_sm_c_lp_do_nm_genes_polarized.log
script c_lp_sm_c_lp_do_nm_genes_polarized.log

java -jar /opt/jvarkit/dist/vcffilterjdk.jar -e 'if(variant.getNAlleles()!=2 || !variant.hasAttribute("AA")) return true; 
final String aa = variant.getAttributeAsString("AA",""); 
if(!variant.getAlleles().get(1).getDisplayString().equalsIgnoreCase(aa)) return true; 
VariantContextBuilder vb=new VariantContextBuilder(variant); 

Allele oldalt = variant.getAlleles().get(1);
Allele oldref = variant.getAlleles().get(0); 
Allele ref= Allele.create(oldalt.getDisplayString(),true); 
Allele alt= Allele.create(oldref.getDisplayString(),false);

vb.alleles(Arrays.asList(ref,alt)); 

List genotypes= new ArrayList<>(); 
for(Genotype g: variant.getGenotypes()) 
  { 
  if(!g.isCalled()) 
  { genotypes.add(g); continue;} 
  GenotypeBuilder gb = new GenotypeBuilder(g); 
  List alleles = new ArrayList<>(); 
  for(Allele a:g.getAlleles()) { 
    if(a.equals(oldalt)) { a=ref;} 
    else if(a.equals(oldref)) { a=alt;} 
    alleles.add(a); 
    } 
  if(g.hasPL()) { 
    int pl[] = g.getPL(); 
    int pl2[] = new int[pl.length]; 
    for(int i=0;i< pl.length;i++) pl2[i]=pl[(pl.length-1)-i]; 
    gb.PL(pl2); 
    } 
  if(g.hasAD()) 
    { int ad[] = g.getAD(); 
    int ad2[] = new int[ad.length]; 
    for(int i=0;i< ad.length;i++) ad2[i]=ad[(ad.length-1)-i];
    gb.AD(ad2); 
  } 
  genotypes.add(gb.alleles(alleles).make()); 
  }

vb.attribute("AF",1.0d - Double.parseDouble(variant.getAttributeAsString("AF",""))); vb.attribute("AC",variant.getGenotypes().stream().flatMap(G->G.getAlleles().stream()).filter(A->A.equals(oldref)).count()); 
vb.genotypes(genotypes); 
return vb.make();' -o c_lp_sm_c_lp_do_nm_genes_polarized.vcf c_lp_sm_c_lp_do_nm_genes_aafilled.vcf

grep -v '#' c_lp_sm_c_lp_do_nm_genes_polarized.vcf | wc -l #1341848

```

#6: Set up SnpEff. I'll be following this manual for all SnpEff configuration purposes: http://snpeff.sourceforge.net/SnpEff_manual.html
##Search for the Lynx pardinus database. 
Search for the Lynx pardinus assembly database in the program's pre-built database. As of June the 6th, 2017, the Lynx pardinus genome isn't included in the snpEff database. A second option would be building our own Lynx pardinus database.
##This was already done by Dani, as he will be running this chunk there is no need to re run it.

```{r Set up SnpEff, eval=FALSE, engine='bash'}

java -jar /opt/snpEff/snpEff.jar databases | grep -i pardinus

```

##Build the Lynx pardinus genome database.
In the end we opt to build our own database since we don't know when they will get themselves to add it. This step should be omitted if the desired database was found in the previous step. In the end it's best to use the Lynx rufus genome as reference if the VCF is polarized based on Lynx rufus, else the polarized SNPs are wrongly annotated.

###Add entry to the config file

```{r Set up SnpEff, eval=FALSE, engine='bash'}

#Originally the config was just in the software folder and I didn't have writing permission. If this is the only config file available, writing permission is required, and when annotating later on, the file should be called using the -c command followed by the path to the file.
#However, in my case I believe the tech group created a copy of the file in my folder after I sent them an e-mail, and this is the one that I was able to edit.

cd /home/dkleinman/
mv snpEff.config $C_PATH #I move the config file that appeared in my folder to a subfolder that I created for snpEff
vi snpEff.config  #initiate the editing process

#Following the manual, I added the following two lines (use :o to start editing -watch out for the current line- and :wq to save and exit):

# Lynx_pardinus
LYPA.23.genome : Iberian lynx #from now on, LYPA.23 is the code for the Lynx pardinus reference genome (in snpEff)

# Lynx_pardinus, detailed annotation (obsolete)
LYPA.23b.genome : Iberian lynx #LYPA.23b is the code for the highly detailed annotation of the Lynx pardinus reference genome (in snpEff)

#In the end it's best to use the Lynx rufus genome as reference if the VCF is polarized based on Lynx rufus, else the polarized SNPs are wrongly annotated.

```

###Create directory and move files

```{r Set up SnpEff, eval=FALSE, engine='bash'}

#First for the regular annotation:
mkdir $S_PATH/data/LYPA.23 #create a directory inside the software's dependencies whose name matches the code
cd $S_PATH/data/LYPA.23

scp /GRUPOS/grupolince/Lyp_annotation_Apr14_final/LYPA23C.all.fix.nr.gff3 $S_PATH/data/LYPA.23/ #copy the annotation file (can be gff or gtf) to the newly created directory. This gff file includes CDS, introns, exons and genes, so it's very basic. A more complex version that Maria created which includes lncRNAs, etc., will be considered as custom annotation.
mv LYPA23C.all.fix.nr.gff3 genes.gff #rename the file as the tutorial indicates

mkdir $S_PATH/data/genomes #create a directory inside the software's dependencies called genomes
cd $S_PATH/data/genomes
scp /home/GRUPOS/grupolince/reference_genomes/lynx_pardinus_genome/lp23.fa $S_PATH/data/genomes #copy the reference genome fasta to the new genomes directory
mv lp23.fa LYPA.23.fa #rename the file so that it matches the code


#In the end this won't be used. The detailed (custom) annotation will be intersected using a bed and the -interval command.
#Second, for the detailed annotation:
mkdir $C_PATH/data/LYPA.23b #create a directory inside the software's dependencies whose name matches the code
cd $C_PATH/data/LYPA.23b

scp /GRUPOS/grupolince/Lyp_annotation_Apr14_final/LYPA23C.CDS.GENE_promoters.GENE_introns.UTRs.ncRNA.lncRNA.lncRNA_introns.lncRNA_promoters.UCNE.intergenic.nr.gff3 $C_PATH/data/LYPA.23b/ #copy the annotation file (can be gff or gtf) to the newly created directory. This gff file is very detailed and includes CDS, introns, exons, genes, and many more.
mv LYPA23C.CDS.GENE_promoters.GENE_introns.UTRs.ncRNA.lncRNA.lncRNA_introns.lncRNA_promoters.UCNE.intergenic.nr.gff3 genes.gff #rename the file as the tutorial indicates

cd $C_PATH/data/genomes
scp LYPA.23.fa LYPA.23b.fa #copy the reference genome and rename it so that it also matches the detailed annotation entry

```

###Build the database

```{r Set up SnpEff, eval=FALSE, engine='bash'}

cd $C_PATH
screen -S build_snpEff_db #open a dettachable screen in case the database building takes too long
script build_snpEff_db.txt #initiate the log file

S_PATH=/opt/snpEff #redefine the variable, since we're inside a script
C_PATH=/home/dkleinman/datos/snpEff #redefine the variable, since we're inside a script

cd $S_PATH
java -jar snpEff.jar build -gff3 -v LYPA.23 -c $C_PATH/snpEff.config -dataDir $C_PATH/data #build the database. Use the -gff3 command for gff files and -gtf22 for gtf files. Use -v for verbose (expanded information on the processes and the warnings/errors that may appear). Use -c to indicate the path to my own config file. Then use -dataDir to override the data directory from the config file (by default the software thinks that the data folder with the genome and the genes files is located where config is, so it's necessary to give it the correct path).

ctrl + D #terminate the script
ctrl + D #terminate the screen

#The following is obsolete since I fixed the original code.
scp -r $S_PATH/data $C_PATH #afterwards I realize anyone can access the data folder so I copy it to my own folder and then I remove the stuff I created inside the original data folder
cd $S_PATH/data
rm -r LYPA.23/
rm -r genomes/

```

### Tutorial annotation
Annotate one of the examples that comes with the software

```{r Tutorial annotation, eval=FALSE, engine='bash'}

java -Xmx16g -jar $S_PATH/snpEff.jar GRCh37.75 -s $O_PATH/toys/test.chr22.ann $S_PATH/examples/test.chr22.vcf > $O_PATH/toys/test.chr22.ann.vcf

```

##Build the Lynx rufus genome database.
In the end we opt to build our own database since we don't know when they will get themselves to add it. This step should be omitted if the desired database was found in the previous step. Annotating against the Lynx rufus genome is the correct option if the VCF has been polarized based on that genome.

###Add entry to the config file

```{r Set up SnpEff, eval=FALSE, engine='bash'}

#Originally the config was just in the software folder and I didn't have writing permission. If this is the only config file available, writing permission is required, and when annotating later on, the file should be called using the -c command followed by the path to the file.
#However, in my case I believe the tech group created a copy of the file in my folder after I sent them an e-mail, and this is the one that I was able to edit.

cd /home/dkleinman/
mv snpEff.config $C_PATH #I move the config file that appeared in my folder to a subfolder that I created for snpEff
vi snpEff.config  #initiate the editing process

#Following the manual, I added the following two lines (use :o to start editing -watch out for the current line- and :wq to save and exit):

# Lynx_rufus
LYRU.23.genome : Bobcat #from now on, LYRU.23 is the code for the Lynx rufus c_lr_zz_0001_recal1.fa genome (in snpEff)

```

###Create directory and move files

```{r Set up SnpEff, eval=FALSE, engine='bash'}

#First for the regular annotation:
mkdir $C_PATH/data/LYRU.23 #create a directory inside the software's dependencies whose name matches the code
cd $C_PATH/data/LYRU.23

scp /GRUPOS/grupolince/Lyp_annotation_Apr14_final/LYPA23C.all.fix.nr.gff3 $C_PATH/data/LYRU.23/ #copy the annotation file (can be gff or gtf) to the newly created directory. This gff file includes CDS, introns, exons and genes, so it's very basic. A more complex version that Maria created which includes lncRNAs, etc., will be considered as custom annotation.
mv LYPA23C.all.fix.nr.gff3 genes.gff #rename the file as the tutorial indicates

mkdir $C_PATH/data/genomes #create a directory inside the software's dependencies called genomes
cd $C_PATH/data/genomes
scp /home/GRUPOS/grupolince/reference_genomes/lynx_rufus_genome/c_lr_zz_0001_recal1.fa $C_PATH/data/genomes #copy the reference genome fasta to the new genomes directory
mv c_lr_zz_0001_recal1.fa LYRU.23.fa #rename the file so that it matches the code

```

###Build the database

```{r Set up SnpEff, eval=FALSE, engine='bash'}

cd $C_PATH
screen -S build_LYRU_snpEff_db #open a dettachable screen in case the database building takes too long
script build_LYRU_snpEff_db.txt #initiate the log file

S_PATH=/opt/snpEff #redefine the variable, since we're inside a script
C_PATH=/home/dkleinman/datos/snpEff #redefine the variable, since we're inside a script

cd $S_PATH
java -jar snpEff.jar build -gff3 -v LYRU.23 -c $C_PATH/snpEff.config -dataDir $C_PATH/data #build the database. Use the -gff3 command for gff files and -gtf22 for gtf files. Use -v for verbose (expanded information on the processes and the warnings/errors that may appear). Use -c to indicate the path to my own config file. Then use -dataDir to override the data directory from the config file (by default the software thinks that the data folder with the genome and the genes files is located where config is, so it's necessary to give it the correct path).

ctrl + D #terminate the script
ctrl + D #terminate the screen

```

#7: Annotate using SnpEff.
##Create custom annotation bed file
Dani will run this chunk
```{r Annotate using SnpEff, eval=FALSE, engine='bash'}

#Create bed file with custom annotations based on the gff3 that María compiled.
cut -d$'\t' -f1,3,4,5 /GRUPOS/grupolince/Lyp_annotation_Apr14_final/LYPA23C.CDS.GENE_promoters.GENE_introns.UTRs.ncRNA.lncRNA.lncRNA_introns.lncRNA_promoters.UCNE.intergenic.nr.gff3 | awk '{printf ("%s\t%s\t%s\t%s\n", $1, $3, $4, $2)}' > /home/dkleinman/datos/snpEff/data/LYPA23C.CDS.GENE_promoters.GENE_introns.UTRs.ncRNA.lncRNA.lncRNA_introns.lncRNA_promoters.UCNE.intergenic.nr.bed

```

##Annotate the VCF with custom annotation

###For Lynx pardinus. Monomorphic positions (incl. substitutions) are NOT polarized. Grab substitutions from the joint dataset.

```{r Annotate using SnpEff, eval=FALSE, engine='bash'}

#For all positions:

cd $V_PATH/annotation/
screen -S c_lp_sm_c_lp_do_nm_genes_polarized.lr_ann.log #open a dettachable screen in case the test takes too long
script c_lp_sm_c_lp_do_nm_genes_polarized.lr_ann.log #initiate the log file

S_PATH=/opt/snpEff #software path
C_PATH=/home/dkleinman/datos/snpEff #config file path
O_PATH=/home/dkleinman/datos/snpEff #output path
I_PATH=/home/GRUPOS/grupolince/immunocapture/prueba_highdiv #immunocapture path
V_PATH=/GRUPOS/grupolince/lynx_genomes_5x/VCFs_Dani/nmVCFs #VCFs path

cd $O_PATH #run this code from the directory where the config is located.
java -Xmx16g -jar $S_PATH/snpEff.jar LYRU.23 -v -s $V_PATH/annotation/c_lp_sm_c_lp_do_nm_genes_polarized.lr_ann.html -csvStats $V_PATH/annotation/c_lp_sm_c_lp_do_nm_genes_polarized.lr_ann.csv -interval $C_PATH/data/LYPA23C.CDS.GENE_promoters.GENE_introns.UTRs.ncRNA.lncRNA.lncRNA_introns.lncRNA_promoters.UCNE.intergenic.nr.bed $V_PATH/c_lp_sm_c_lp_do_nm_genes_polarized.vcf > $V_PATH/annotation/c_lp_sm_c_lp_do_nm_genes_polarized.lr_ann.vcf #run this code from the directory where the config is located.

cd $V_PATH/annotation/
grep -v '#' c_lp_sm_c_lp_do_nm_genes_polarized.lr_ann.vcf | wc -l #1341848

```


#8: Filter the annotated VCF. Subset the VCF files in order to keep only good quality biallelic SNP variants.
##For Lynx pardinus.
Remember to use mu cnv_file to filter and keep only songle copy genes
Evaluate tjhe possibility to pipe filters and get one VCF directy
```{r Filter the annotated VCF, eval=FALSE, engine='bash'}

cd $V_PATH/annotation
screen -S c_lp_sm_c_lp_do_nm_genes_polarized_filtered.lr_ann.log
script c_lp_sm_c_lp_do_nm_genes_polarized_filtered.lr_ann.log

V_PATH=/GRUPOS/grupolince/lynx_genomes_5x/VCFs_Dani/nmVCFs #VCFs path
REF=/home/GRUPOS/grupolince/reference_genomes/lynx_pardinus_genome/lp23.fa #path to reference genome (first eight scaffolds only)
GATK=/opt/GATK-3.7/GenomeAnalysisTK.jar #GATK software path
BCF=/opt/bcftools-1.6/bcftools #BCFtools software path


#Remove repetitive regions and those with low mappability or non-single copy:
bedtools subtract -a c_lp_sm_c_lp_do_nm_genes_polarized.lr_ann.vcf -b /home/GRUPOS/grupolince/reference_genomes/lynx_pardinus_genome/repeats_and_lowcomp_no_redundant_mappability.k75_lessthan90.bed -header | !!! ADD MY SINGLE COPY BED!!> c_lp_sm_c_lp_do_nm_genes_polarized_filtered1.lr_ann.vcf

!!! ADD MY SINGLE COPY BED!!
grep -v '#' c_lp_sm_c_lp_do_nm_genes_polarized_filtered1.lr_ann.vcf | wc -l #645623


#During this step, all INDELs as well as all multiallelic and bad quality SNPs will be dropped from the respective VCFs.
java -XX:MaxMetaspaceSize=1g -XX:+UseG1GC -XX:+UseStringDeduplication -Xms16g -Xmx32g -jar $GATK \
-T SelectVariants \
-selectType SNP \
-restrictAllelesTo BIALLELIC \
-R $REF \
-V c_lp_sm_c_lp_do_nm_genes_polarized_filtered1.lr_ann.vcf \
-o c_lp_sm_c_lp_do_nm_genes_polarized_filtered2.lr_ann.vcf

grep -v '#' c_lp_sm_c_lp_do_nm_genes_polarized_filtered2.lr_ann.vcf | wc -l #497025


#Remove polarized ANC fixed variants (AF=0 or AF=1) or those that weren't polarizable (AA different from either REF or ALT):
$BCF view -e '(INFO/AA!=REF & INFO/AA!=ALT) | INFO/AF=0.00 | INFO/AF=1.00' c_lp_sm_c_lp_do_nm_genes_polarized_filtered2.lr_ann.vcf > c_lp_sm_c_lp_do_nm_genes_polarized_filtered3.lr_ann.vcf

grep -v '#' c_lp_sm_c_lp_do_nm_genes_polarized_filtered3.lr_ann.vcf | wc -l #475783

#Apply GATK's recommended filters, and then some.     
java -XX:MaxMetaspaceSize=1g -XX:+UseG1GC -XX:+UseStringDeduplication -Xms16g -Xmx32g -jar $GATK \
-T SelectVariants \
-select "QUAL >= 30 && QD >= 2.0 && FS <= 60.0 && MQ >= 40.0 && MQRankSum >= -12.5 && ReadPosRankSum >= -8.0" \
-R $REF \
-V c_lp_sm_c_lp_do_nm_genes_polarized_filtered3.lr_ann.vcf \
-o c_lp_sm_c_lp_do_nm_genes_polarized_filtered4.lr_ann.vcf

grep -v '#' c_lp_sm_c_lp_do_nm_genes_polarized_filtered4.lr_ann.vcf | wc -l #446094

#First, for each species exclude those positions that have more than 20% missing genotypes
Aquí me merece la pena explorar qué % de missingness me puedo permitir filtrar (sobre todo con las históricas)

$BCF filter -e " F_MISSING > 0.2" -Ov -o c_lp_sm_c_lp_do_nm_genes_polarized_filtered5.lr_ann.vcf c_lp_sm_c_lp_do_nm_genes_polarized_filtered4.lr_ann.vcf

grep -v '#' c_lp_sm_c_lp_do_nm_genes_polarized_filtered5.lr_ann.vcf | wc -l #402692

```

#9: Obtain per population VCFs.
##Split the VCF into per population VCFs. Generate a VCF for each population.
###For Lynx pardinus and Lynx lynx.

```{r Obtain per population VCFs, eval=FALSE, engine='bash'}

cd $V_PATH/annotation/
screen -S perpop_nm_genes.lr_ann.log
script perpop_nm_genes.lr_ann.log

cd $B_PATH/BAM_genes_5x
declare SPECIES=$(ls {*_lp_*,*_ll_*}_recal_round-1.genes-nm.bam | cut -c3-4 | sort | uniq)
cd $V_PATH/annotation/
for i in ${SPECIES[@]}
  do
  echo "${i}"
  cd $B_PATH/BAM_genes_5x
  declare POP=$(ls c_{lp_sm*,lp_do*,ll_ki*,ll_po*,ll_no*}_recal_round-1.genes-nm.bam | cut -c1-7 | uniq | grep "${i}")
  cd $V_PATH/annotation/
  for j in ${POP[@]}
    do
    echo "${j}"
    rm "${j}"_pop_list_to_remove.txt
    $BCF query -l $V_PATH/annotation/c_"${i}"*nm_genes_polarized_filtered5.lr_ann.vcf | grep "${j}" > "${j}"_pop_list_to_remove.txt
    cat "${j}"_pop_list_to_remove.txt
    mkdir "${j}"_perpop_nm_genes
    java -XX:MaxMetaspaceSize=1g -XX:+UseG1GC -XX:+UseStringDeduplication -Xms16g -Xmx32g -jar $GATK \
    -T SelectVariants \
    -R $REF \
    -V $V_PATH/annotation/c_"${i}"*nm_genes_polarized_filtered5.lr_ann.vcf \
    -o $V_PATH/annotation/"${j}"_perpop_nm_genes/"${j}"_perpop_nm_genes.lr_ann.vcf \
    -env \
    --sample_file "${j}"_pop_list_to_remove.txt
    rm "${j}"_pop_list_to_remove.txt
    grep -v '#' $V_PATH/annotation/"${j}"_perpop_nm_genes/"${j}"_perpop_nm_genes.lr_ann.vcf | wc -l #
    done
  done

cd $V_PATH/annotation/
grep -v '#' c_ll_no_perpop_nm_genes/c_ll_no_perpop_nm_genes.lr_ann.vcf | wc -l #
grep -v '#' c_ll_po_perpop_nm_genes/c_ll_po_perpop_nm_genes.lr_ann.vcf | wc -l #
grep -v '#' c_ll_ki_perpop_nm_genes/c_ll_ki_perpop_nm_genes.lr_ann.vcf | wc -l #
grep -v '#' c_lp_do_perpop_nm_genes/c_lp_do_perpop_nm_genes.lr_ann.vcf | wc -l #259960
grep -v '#' c_lp_sm_perpop_nm_genes/c_lp_sm_perpop_nm_genes.lr_ann.vcf | wc -l #374157

```

#10: Split the population VCFs into per individual VCFs. Generate a VCF for each individual.
##For Lynx pardinus and Lynx lynx.
Evaluate wheter I need this chunk
```{r Split the population VCFs into per individual VCFs, eval=FALSE, engine='bash'}

#Generate individual VCFs from the per species VCFs:
cd $V_PATH/annotation
screen -S individuals_nm_genes.lr_ann.log
script individuals_nm_genes.lr_ann.log

#For each individual in the sm, do, ki, po & no populations, subset its variants from the respective population VCF:
cd $B_PATH/BAM_genes_5x
declare POP=$(ls c_{lp_sm*,lp_do*,ll_ki*,ll_po*,ll_no*}_recal_round-1.genes-nm.bam | cut -c1-7 | uniq | grep "${i}")
for i in ${POP[@]}
  do
  echo "${i}"
  mkdir $V_PATH/annotation/"${i}"_individuals_nm_genes
  declare INDIVIDUALS=$(ls "${i}"*_recal_round-1.genes-nm.bam | cut -c1-12 | uniq)
  for j in ${INDIVIDUALS[@]}
    do
    echo "${j}"
    ID=$(echo "${j}")
    java -XX:MaxMetaspaceSize=1g -XX:+UseG1GC -XX:+UseStringDeduplication -Xms16g -Xmx32g -jar $GATK \
    -T SelectVariants \
    -R $REF \
    -V $V_PATH/annotation/"${i}"_perpop_nm_genes/"${i}"_perpop_nm_genes.lr_ann.vcf \
    -o $V_PATH/annotation/"${i}"_individuals_nm_genes/"${j}"_individual_nm_genes.lr_ann.vcf \
    -env \
    -sn $ID
    done
  done

```

#11: Get annotation statistics.
##At the individual level.
###Heterozygotes and derived/derived homozygotes.

```{r Get annotation statistics, eval=FALSE, engine='bash'}

S_PATH=/opt/snpEff #software path
C_PATH=/home/dkleinman/datos/snpEff #config file path
O_PATH=/home/dkleinman/datos/snpEff #output path
I_PATH=/home/GRUPOS/grupolince/immunocapture/prueba_highdiv #immunocapture path
V_PATH=/GRUPOS/grupolince/lynx_genomes_5x/VCFs_Dani/nmVCFs #VCFs path
G_PATH=/GRUPOS/grupolince/lynx_genomes_5x/gVCFs #gVCFs path
B_PATH=/home/GRUPOS/grupolince/lynx_genomes_5x/BAM_files_final #BAM files path
REF=/home/GRUPOS/grupolince/reference_genomes/lynx_pardinus_genome/lp23.fa #path to reference genome
GATK=/opt/GATK-3.7/GenomeAnalysisTK.jar #GATK software path
BCF=/opt/bcftools-1.6/bcftools #BCFtools software path


#For Lynx pardinus and Lynx lynx:

cd $V_PATH/annotation/
screen -S snpeff_individual_summary_nm_genes.lr_ann.log
script snpeff_individual_summary_nm_genes.lr_ann.log

rm snpeff_individual_summary_nm_genes.lr_ann.txt
echo -e "species\tpopulation\tsample\ttotal_V\tintergenic_V\tintronic_V\tcoding_V\tsynonymous_V\tsynonymous_A\tmissense_V\tmissense_A\tnonsense_V\tnonsense_A\tUCNE_V\tmissense/synonymous_V\tmissense/synonymous_A\tsynonymous/intronic_V\tmissense/intronic_V" > snpeff_individual_summary_nm_genes.lr_ann.txt
INDLIST=($(ls `find . -name '*individual_nm_genes.lr_ann.vcf' -print`))
for i in "${INDLIST[@]}"
  do
  echo "${i}"
  ind=$(echo "${i}" | awk -F'[/]' '{print $3}')
  echo "${ind}"
  SPECIES=$(echo "${ind}" | cut -c3-4)
  POPULATION=$(echo "${ind}" | cut -c6-7)
  SAMPLE=$(echo "${ind}" | cut -c9-12)
  TOTAL_V=$(grep -v '#' ${i} | wc -l)
  INTERGENIC_V=$(grep 'intergenic' ${i} | wc -l)
  INTRONIC_V=$(grep 'intron_variant' ${i} | wc -l)
  CODING_V=$(grep 'CDS' ${i} | wc -l)
  SYNONYMOUS_V=$(grep 'synonymous_variant' ${i} | wc -l)
  SYNONYMOUS_A=$(grep 'synonymous_variant' ${i} | cut -f8 | cut -d';' -f2 | cut -d'=' -f2 | paste -sd+ | bc)
  MISSENSE_V=$(grep 'missense_variant' ${i} | wc -l)
  MISSENSE_A=$(grep 'missense_variant' ${i} | cut -f8 | cut -d';' -f2 | cut -d'=' -f2 | paste -sd+ | bc)
  NONSENSE_V=$(grep -e 'lost' -e 'gained' ${i} | wc -l)
  NONSENSE_A=$(grep -e 'lost' -e 'gained' ${i} | cut -f8 | cut -d';' -f2 | cut -d'=' -f2 | paste -sd+ | bc)
  UCNE_V=$(grep 'UCNE' ${i} | wc -l)
  MISSENSE_SYNONYMOUS_V=$(echo "scale=3; $MISSENSE_V/$SYNONYMOUS_V" | bc)
  MISSENSE_SYNONYMOUS_A=$(echo "scale=3; $MISSENSE_A/$SYNONYMOUS_A" | bc)
  SYNONYMOUS_INTRONIC_V=$(echo "scale=3; $SYNONYMOUS_V/$INTRONIC_V" | bc)
  MISSENSE_INTRONIC_V=$(echo "scale=3; $MISSENSE_V/$INTRONIC_V" | bc)
  echo -e "$SPECIES\t$POPULATION\t$SAMPLE\t$TOTAL_V\t$INTERGENIC_V\t$INTRONIC_V\t$CODING_V\t$SYNONYMOUS_V\t$SYNONYMOUS_A\t$MISSENSE_V\t$MISSENSE_A\t$NONSENSE_V\t$NONSENSE_A\t$UCNE_V\t$MISSENSE_SYNONYMOUS_V\t$MISSENSE_SYNONYMOUS_A\t$SYNONYMOUS_INTRONIC_V\t$MISSENSE_INTRONIC_V" >> $V_PATH/annotation/snpeff_individual_summary_nm_genes.lr_ann.txt
  done

scp dkleinman@genomics-b.ebd.csic.es:/GRUPOS/grupolince/lynx_genomes_5x/VCFs_Dani/nmVCFs/annotation/snpeff_individual_summary_nm_genes.lr_ann.txt /Users/Dani/ownCloud/backup/g-w_analysis/genetic_load/snpeff_summary_ratios/sep_calling

```
